Sparse Spectral Methods and P-Finite Element Methods on Disk-Splices and Rhombi - Part I: Zero Boundary Conditions

### 2018-12-11 papers/halfdisk/halfdisk.tex
Further edits to paper

\newcommand{\Pnk}{P_{n,k}}
\newcommand{\Pnkab}{P_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\Pmjab}{P_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bigPab}{\bigP^{(a,b)}}
\newcommand{\Dnt}{D^T_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\Pii}{P^{(1,1)}}
\newcommand{\Poo}{P^{(0,0)}}
\newcommand{\Poi}{P^{(0,1)}}
\newcommand{\Pio}{P^{(1,0)}}
\newcommand{\dx}{\frac{\partial}{\partial x}}
\newcommand{\dy}{\frac{\partial}{\partial y}}

@@

\begin{align}
\alpha &:= 0 \\
\beta &:= 1 \\
\gamma &:= -1 \\
\delta &:= -c = 1 \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= x^a \: (1-x^2)^b \\
w_2(x) &:= (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{align}

@@

This provides us with a family of orthogonal polynomial spaces on the half disk.

@@

We can develop a quadrature rule on \(\Omega\) in order to evaluate (40) and obtain the coefficients vector \(\mathbf{f}\) as follows.

@@

@@

For general \(a,b\) and general function \(f(x,y) = \mathbf{f}^T \: \bigPab(x,y)\), there exist Banded-Block-Banded matrix operators \(D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}\) such that:
\begin{align}
\dx[f(x,y)] &= \bigP^{(a+1,b+1)}(x,y)^T \: D_x^{(a,b)} \: \mathbf{f}, \\
\dy[f(x,y)] &= \bigP^{(a,b+1)}(x,y)^T \: D_y^{(a,b)} \: \mathbf{f}, \\
\dx[\Wab(x,y) \: f(x,y)] &= W^{(a-1,b-1)}(x,y) \: \bigP^{(a-1,b-1)}(x,y)^T \: W_x^{(a,b)} \: \mathbf{f}, \\
\dy[\Wab(x,y) \: f(x,y)] &= W^{(a,b-1)}(x,y) \: \bigP^{(a,b-1)}(x,y)^T \: W_y^{(a,b)} \: \mathbf{f}.
\end{align}
The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomials families' derivatives, for example the Jacobi polynomials, as seen in the DLMF \cite{DLMFDerivatives}.
Further, we note that there exist Banded-Block-Banded matrix operators \(T^{(1,0)\to(0,0)}\) and \(T^{(0,1)\to(1,1)}\) such that:
\begin{align}
x \bigP^{(1,0)}(x,y) &= \Big(T^{(1,0)\to(0,0)} \Big)^T \: \bigP^{(0,0)}(x,y) \\
\bigP^{(0,1)}(x,y) &= \Big(T^{(0,1)\to(1,1)} \Big)^T \: \bigP^{(1,1)}(x,y)
\end{align}
Then, the operator \(\Delta\) is given by
\begin{align}
    \Delta := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T^{(1,0)\to(0,0)} \: W_y^{(1,1)}
\end{align}

%
\section{P-finite element methods using sparse operators}
Consider the 2D Dirichlet problem on a disk-splice or rhobus $\Omega$:
\begin{align}
	\begin{cases}
                \Delta u(x,y) = f(x,y) \quad \text{in } \Omega \\
                u(x,y) = 0 \quad \text{on } \partial \Omega
         \end{cases}
\end{align}
This has the weak formulation for any test function $v$
\begin{align}
	L(v) := \int \int_\Omega f \: v \: dy \: dx = - \int \int_\Omega \nabla u \cdot \nabla v \: dy \: dx =: a(u,v).
\end{align}

We can choose our finite dimensional space $V$ to

@@

\subsection{Heat Equation}

Fix \(a, b = 1,1\), and consider
\begin{align}
\frac{\partial u}{\partial t}(x,y) = \Delta(\Wii \: u)(x,y) \quad \text{in } \Omega
\end{align}

>>>>

\newcommand{\Pnk}{P_{n,k}}
\newcommand{\Pnkab}{P_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\Pmjab}{P_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bigPab}{\bigP^{(a,b)}}
\newcommand{\Dnt}{D^T_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\Pii}{P^{(1,1)}}
\newcommand{\Poo}{P^{(0,0)}}
\newcommand{\Poi}{P^{(0,1)}}
\newcommand{\Pio}{P^{(1,0)}}
\newcommand{\dx}{\frac{\partial}{\partial x}}
\newcommand{\dy}{\frac{\partial}{\partial y}}
\newcommand{\element}{\tau}
\newcommand{\refelement}{\hat{\tau}}
\newcommand{\FEset}{\mathcal{T}}

@@

\begin{align}
\begin{cases}
\alpha &:= 0 \\
\beta &:= 1 \\
\gamma &:= -1 \\
\delta &:= -\gamma = 1 \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= x^a \: (1-x^2)^b \\
w_2(x) &:= (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align}

@@

This provides us with a family of orthogonal polynomial spaces on the half disk. Note that we can obtain OP families on other disk-splices and rhombi by appropriately choosing the functions and constants in (2). \bstodo{Should I add in the rhombi / other disk splices formulations here?}

@@

We can develop a quadrature rule on \(\Omega\) in order to evaluate (40) and obtain the coefficients vector \(\mathbf{f}\) as follows. \bstodo{Should I add in the rhombi / other disk splices formulations here?}

@@

Note that the sparsity of the Jacobi matrices (in particular the sparsity of the sub-blocks) comes from the natural sparsity of the three-term recurrences of the 1D OPs. In other words, sparsity is not dependent on the domain.

@@

For general \(a,b\) and general function \(f(x,y) = \mathbf{f}^T \: \bigPab(x,y)\), there exist Banded-Block-Banded matrix operators \(D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}\) such that:
\begin{align}
\dx[f(x,y)] &= \bigP^{(a+1,b+1)}(x,y)^T \: D_x^{(a,b)} \: \mathbf{f}, \\
\dy[f(x,y)] &= \bigP^{(a,b+1)}(x,y)^T \: D_y^{(a,b)} \: \mathbf{f}, \\
\dx[\Wab(x,y) \: f(x,y)] &= W^{(a-1,b-1)}(x,y) \: \bigP^{(a-1,b-1)}(x,y)^T \: W_x^{(a,b)} \: \mathbf{f}, \\
\dy[\Wab(x,y) \: f(x,y)] &= W^{(a,b-1)}(x,y) \: \bigP^{(a,b-1)}(x,y)^T \: W_y^{(a,b)} \: \mathbf{f}.
\end{align}
The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomials families' derivatives, for example the Jacobi polynomials, as seen in the DLMF \cite{DLMFDerivatives}.
Further, we note that there exist Banded-Block-Banded matrix operators \(T^{(1,0)\to(0,0)}\) and \(T^{(0,1)\to(1,1)}\) such that:
\begin{align}
x \bigP^{(1,0)}(x,y) &= \Big(T^{(1,0)\to(0,0)} \Big)^T \: \bigP^{(0,0)}(x,y) \\
\bigP^{(0,1)}(x,y) &= \Big(T^{(0,1)\to(1,1)} \Big)^T \: \bigP^{(1,1)}(x,y)
\end{align}
Then, the operator \(\Delta\) is given by
\begin{align}
    \Delta := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T^{(1,0)\to(0,0)} \: W_y^{(1,1)}.
\end{align}
The sparsity of the differential operators can be shown using a simply integration by parts argument, again showing that the sparsity for differential operators is not dependent on the domain.

%
\section{P-finite element methods using sparse operators}
\bstodoinline{Need to discuss this}
We follow the method of \cite{beuchler2006new}. Consider the 2D Dirichlet problem on a domain $\Omega$:
\begin{align}
	\begin{cases}
         - \Delta u(x,y) = f(x,y) \quad \text{in } \Omega \\
         u = 0 \quad \text{on } \partial \Omega
         \end{cases}
\end{align}
This has the weak formulation for any test function $v \in V := H_0^1(\Omega) = \{v \in H^1(\Omega) \quad | \quad v|_{\partial \Omega} = 0 \}$,
\begin{align}
	L(v) := \int_\Omega f \: v \: d\mathbf{x} = \int_\Omega \nabla u \cdot \nabla v \: d\mathbf{x} =: a(u,v).
\end{align}

Let $\FEset$ be the set of elements $\element$ that make up our finite element discretisation of the domain, where each $\element$ is a rhombus for example. We can choose our finite dimensional space $V_h = \{v_h \in V \quad | \quad deg(v_h)|_\element \le p \: \forall \element \in \FEset\}$.

We seek $u_h \in V_h$ s.t.
\begin{align}
	L(v_h) = a(u_h,v_h) \quad \forall \: v_h \in V_h.
\end{align}

We can choose a basis for $V_h$ by using the orthogonal polynomials 
\begin{align}
\bigP(x,y) := \begin{bmatrix}
		\bigP_0(x,y) \\
		\hline
		\bigP_1(x,y) \\
		\hline
		\bigP_2(x,y) \\
		\hline
		\vdots \\
		\hline
		\bigP_p(x,y)
	\end{bmatrix},
\quad
\bigP_n(x,y) := \begin{bmatrix}
		P_{n,0}(x,y) \\
		\vdots \\
		P_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,p,
\end{align}
Appropriately relabelling these OPs as the set $\{\phi_1,\dots,\phi_N\}$, we can define the global stiffness matrix $A$ by $A_{ij} =  a(\phi_i, \phi_j)$. The global stiffness matrix can be computed using the element stiffness matrices $A^\element = \int_\element \nabla \bigP \cdot \nabla \bigP \: d\mathbf{x}, \: \element \in \FEset$. Then we have the linear system for our finite element problem:
\begin{align}
A\mathbf{u} = \mathbf{f}
\end{align}
where the vectors $\mathbf{u}, \mathbf{f}$ are the coefficient vectors for expansion of the solution $u_h$ and for the function $f$ respectively.

Using the quadrature rule we have formulated for integrals on a rhombus element $\element$, and the sparse operator matrices for divergence (partial derivatives), we obtain a sparse global stiffness matrix.

@@

\subsection{Heat Equation}
\bstodo{Struggling to get plotting working}
Fix \(a, b = 1,1\), and consider
\begin{align}
\frac{\partial u}{\partial t}(x,y) = \Delta(\Wii \: u)(x,y) \quad \text{in } \Omega
\end{align}
